Degrees Draft Syllabus

### 2023-03-24 Draft Syllabus.Rmd
update to lesson 3, correct tidy data location
Lessons 1-3 final draft

## Introduction

![Welcome!](./Introduction/welcome.jpg) <br/>
Welcome to the 2023 #DegreesNYC Introduction to Data Analysis Course! As you know, #DegreesNYC is on a mission to radically transform the NYC public school system at all levels. Collective impact and youth engagement are pillars of #DegreesNYC's mission and are the primary tools that the organization uses to further its mission. The folks at #DegreesNYC, however, are aware that the policy space in which they exist values quantitative evidence highly. The Data Fellowship is designed to democratize quantitative expertise, giving power to those who are closest to the issues #DegreesNYC addresses and most impacted by the current state of NYC education: students. This course is meant to give you an introduction to some of the skills you will need to find, manipulate, and analyze data using R. Our hope is that this will not only be enriching for you personally, but will better position you to make policy arguments in the future.

During the course you are going to work primarily (exclusively?) in R and much of the work done using R will be done with R Markdown. Though there are many other languages used for data analysis (including Python, SAS, Stata, and SQL), you will not explore those. If you'd like to know more about one of them feel free to reach out to me and I am happy to share what I know or help point you to a starting place if I am inexperienced in the language myself. And it is okay to be inexperienced! Don't worry if you are not familiar with the language or with coding at all, this class and the Codecademy course that will supplement it are meant for beginners. If you do have experience with R and feel that the course is moving too slowly or that you are not being sufficiently challenged, you encourage you to 1) Add complexity to the examples and challenges presented in class or on Codecademy, 2) Visit some of the pages provided in the enrichment section of this syllabus, 3) read ahead in the syllabus and work on something that is at your level, or 4) focus extra time on your project.

### About the instructor

I am a Data Analytics Developer at Mathematica Policy Research. Before going to Mathematica I was a PhD student at Teachers College, Columbia University in the Economics and Education Program (I am currently on a leave of absence). Before my time at TC, I worked for the Research Alliance for New York City Schools. This is where my relationship with #DegreesNYC began. I have experience using SAS, SQL, Stata, R, and Python (as well as tools like Excel), but I am most well versed in R and Stata.

### Plan for the course

@@

### Syllabus

This syllabus will be the repository for everything course-related. Information about lessons, the project, and enrichment is all contained here. The table of contents in the top left has links to the content for units 1 and 2, the introduction, and enrichment activities. Within the Unit 1 header there are tabs for each of the XXXX lessons that you will work through during our time together. These lessons each also have tabs for the topics you will cover, the lesson itself, and any helpful links that you might want to explore. This document will be hosted publically on RPubs, but it is a living document. If anything changes or if there are problems with the URL leading to the syllabus, I will let you know as soon as possible.

@@

## Lessons {.tabset .tabset-fade .tabset-dropdown}

### Lesson 1 - An introduction to R (what, why, how) {.tabset .tabset-fade .tabset-pills}

#### Topics Covered

@@

#### Lesson

##### **What is R?**

- Fully realized programming language including loops, conditionals, and user-defined functions.
- Strong data storage, manipulation, and visualization capabilities.
- Suite of statistical packages and built-in functions.

##### **Why use R?**

![R is a very common language in data science](./Lesson 1/r_prevalence.png)
R is one of the most common languages for data science and data analysis. This is because, with experience, one can take a project from beginning to end only using R. You can download data; manipulate it by creating new variables, dropping missing observations, etc.; perform statistical analysis; and visualize and output results. R, like Stata and SAS, has built-in statistical analysis functionality, but unlike those two languages it is 100% free and open source. There is a robust range of functions in base R, but the R community is very strong and has built a host of tools that make doing data analysis and visualization much easier than it is in base R (you will be working with one such suite of tools, the tidyverse/dplyr, in this class). If you are hoping to get a job doing analytic work, if you want advanced education in a social science field, if you want an introduction to coding/computer science, or if you are just looking to better understand quantitative analysis, R is a good place to begin.

##### **How do I install R?**

@@

##### **Class tools**

As noted in the introduction, this class will use Codecademy to aid in learning basic coding skills in R and the basic R workflow. You will need a Codecademy account to retain your progress between sessions. You can register for one [here](https://www.codecademy.com/)
<!-- Should I add in Git here? -->
Another tool you will use throughout the class and one you will also be using in your Codecademy work is a suite of R packages called the Tidyverse. The Tidyverse is designed to make working with data easier and is commonly used in many organizations (you use it at Mathematica, for example). There are eight major packages in the Tidyverse as well as many other specialised packages, but the following are important for this class:

1. ggplot2 - For visualizing data (i.e. creating graphs, and charts).
2. dplyr - For cleaning data and provides access to the pipe (%>%) which you will make use of for many of your data cleaning tasks in this class.
3. readr - For reading data sets with simple syntax.

To install the Tidyverse, open R or RStudio and type *install.packages("tidyverse")* into the console.

@@

##### **Getting help in R**

@@

1. If you are ever unsure what a function does you can type ?functionname into the console and a help file will pop up in the bottom right corner of your screen that has useful information about the function.

@@

2. If you are curious about an entire package you can type vignette(packagename) for more information about the package.

@@

3. If you need help resolving a more complex problem, try Google or another search engine first. There are many people working in R and many sites which aggregate responses to frequently asked questions. Stackoverflow is one that is well-known in the coding community and that I personally use regularly while at work.

#### Helpful Links

@@

### Lesson 2 - Variables, data types, and operators {.tabset .tabset-fade .tabset-pills}

#### Topics Covered

@@

#### Lesson

##### **Printing output**

@@

Another way that is often used and more flexible than just writing in the console is to wrap the item you want to show in the *print* function.

@@

##### **An introduction to variables**

Variables in R are convenient ways to keep track of values. There are times when you will want to perform an operation. Let's work with the example of addition. You could do the following to add 4 twice:

@@

Variables allow us to work with values dynamically and without the struggle of trying to remember all of the values that you are working with individually. Keeping the math theme, suppose you wanted to implement the quadratic formula which is written:<br/>
![](./Lesson 2/quadratic_formula.png)<br/>

@@

##### **Assigning variables in R**

@@

There is nothing wrong with using any of the assignment operators so you should go with the one you are most comfortable with, especially as you learn. The industry standard, however, is to use the left-arrow unless there is good reason. This prevents confusion when you are trying to do a check for equality (==) and may be easier to read than the right-arrow.

##### **Basic data types**

@@

1. Logical
    + True/false data type (Boolean).
2. Numeric
    + All real numbers.
3. Integer
    + Real values without decimal points.
4. Complex
    + Imaginary values
5. Character
    + String values. These are values that contain non-numeric characters (such as letters).
6. Raw
    + Specifies values as raw bytes

@@

###### **Logical**

Data with class logical can only take on the value of true or false. Another name for this type of data is Boolean. In R, TRUE and FALSE are equivalent to 1 and 0 respectively. That means you can do things like add and subtract logical data even though they are not technically numeric.

@@

###### **Numeric**

@@

###### **Character**

@@

The first line of code throws an error, telling us that "1" and "2" are not numeric and therefore cannot be added. The second line of code returns what you expect, 3.

##### **Mathematical operators**

@@

##### **Logical operators**

You are also likely familiar with most of the logical operators. These are things in math which do not return a number, but rather return a value of true or false. Greater than (>), less than (<), greater than or equal to (>=), less than or equal to (<=), equal to (==), and not equal to (!=) are all available in R. Note that you must use 2 equal signs in R when you what to equate two things. One equal sign, as you discussed earlier, is for assigning values to variables.

@@

There are also operators for *or* (|), *and* (&), and *not* (!).

@@

#### Helpful Links

* [You can read more about the data types that you do not cover here.](https://www.programiz.com/r/data-types)
* [More information about operators](https://www.statmethods.net/management/operators.html)

### Lesson 3 - Tidy data and dplyr {.tabset .tabset-fade .tabset-pills}

#### Topics Covered

@@

#### Lesson

##### **Why dplyr?**

@@

##### **%>% and dplyr functions**

###### **%>%**

@@

###### **filter**

@@

###### **slice**

@@

###### **arrange**

@@

###### **select**

@@

###### **rename**

@@

###### **mutate**

@@

###### **relocate**

@@

###### **summarize**

@@

### Lesson 4 - Vectors and lists {.tabset .tabset-fade .tabset-pills}

#### Topics Covered

@@

### Lesson 5 - Conditionals and loops {.tabset .tabset-fade .tabset-pills}

#### Topics Covered

@@

### Lesson 6 - Functions {.tabset .tabset-fade .tabset-pills}

>>>>

# Introduction

![Welcome!](./Introduction/welcome.jpg) <br/>
Welcome to the 2023 #DegreesNYC R Course! As you know, #DegreesNYC is on a mission to radically transform the NYC public school system at all levels. Collective impact and youth engagement are pillars of #DegreesNYC's mission and are the primary tools that the organization uses to further its goals. The folks at #DegreesNYC, however, are aware that the policy space in which they exist values quantitative evidence highly. This course is meant to democratize quantitative expertise, by giving you an introduction to some of the skills you will need to find, manipulate, and analyze data using R. Our hope is that this will not only be enriching for you personally, but will better position you to make policy arguments in the future.

During the course you are going to work primarily in R and much of the work done using R will be done with R Markdown. Though there are many other languages used for data analysis (including Python, SAS, Stata, and SQL), you will not explore those. If you'd like to know more about one of them feel free to reach out to me and I am happy to share what I know or help point you to a starting place if I am inexperienced in the language myself. And it is okay to be inexperienced! Don't worry if you are not familiar with the language or with coding at all; this class and the Codecademy course that will supplement it are meant for beginners. If you do have experience with R and feel that the course is moving too slowly or that you are not being sufficiently challenged, I encourage you to 1) Add complexity to the examples and challenges presented in class or on Codecademy, 2) Visit some of the pages provided in the enrichment section of this syllabus, 3) read ahead in the syllabus and work on something that is at your level, or 4) focus extra time on your project.

## About the instructor

I am a Data Analytics Developer at [Mathematica](https://www.mathematica.org/). Before going to Mathematica, I was a PhD student at Teachers College, Columbia University in the Economics and Education Program. Before that, I worked for the Research Alliance for New York City Schools. This is where my relationship with #DegreesNYC began. I have experience using SAS, SQL, Stata, R, and Python (as well as tools like Excel), but I am most well versed in R and Stata.

## Plan for the course

@@

## Syllabus

This syllabus will be the repository for everything course-related. Information about lessons, the project, and enrichment is all contained here. The table of contents in the top left has links to the content for units 1 and 2, the introduction, and enrichment activities. Within the Unit 1 header there are tabs for each of the XXXX lessons that you will work through during our time together. These lessons each also have tabs for the topics you will cover, the lesson itself, and any helpful links that you might want to explore. This document will be hosted publicly on GitHub, but it is a living document. If anything changes or if there are problems with the URL leading to the syllabus, I will let you know as soon as possible.

@@

# Unit 1 - Learn R {.tabset .tabset-fade .tabset-dropdown}

## Lesson 1 - An introduction to R (what, why, how) {.tabset .tabset-fade .tabset-pills}

### Topics Covered

@@

### Lesson

#### **What is R?**

- Fully realized programming language including loops, conditionals, user-defined functions, and several object-oriented systems.
- Strong data storage, manipulation, and visualization capabilities.
- Suite of statistical packages and built-in functions.

#### **Why use R?**

![R is a very common language in data science](./Lesson 1/r_prevalence.png)
R is one of the most common languages for data science and data analysis. This is because, with experience, one can take a project from beginning to end only using R. You can download data; manipulate it by creating new variables, dropping missing observations, etc.; perform statistical analysis; and visualize and output results. R, like Stata and SAS, has built-in statistical analysis functionality, but unlike those two languages it is 100% free and open source. There is are a wide range of functions in base R, but the R community is very strong and has built a host of tools that make doing data analysis and visualization much easier than it is in base R (you will be working with one such suite of tools, the tidyverse/dplyr, in this class). If you are hoping to get a job doing analytic work, if you want advanced education in a social science field, if you want an introduction to coding/computer science, or if you are just looking to better understand quantitative analysis, R is a good place to begin.

#### **How do I install R?**

@@

#### **Class tools**

As noted in the introduction, this class will use Codecademy to aid in learning basic coding skills in R and the basic R workflow. You will need a Codecademy account to retain your progress between sessions. You can register for one [here](https://www.codecademy.com/)
<!-- Should I add in Git here? -->
Another tool you will use throughout the class and one you will also be using in your Codecademy work is a suite of R packages called the Tidyverse. The Tidyverse is designed to make working with data easier and is commonly used in many organizations (we use it at Mathematica, for example). There are eight major packages in the Tidyverse as well as many other specialized packages, but the following are important for this class:

1. ggplot2 - For visualizing data (i.e. creating graphs, and charts).
2. dplyr - For cleaning data and provides access to the magrittr pipe (`%>%`) which you can make use of for many of your data cleaning tasks in this class (although R now has a built-in pipe: `|>`).
3. readr - For reading data sets with simple syntax.

To install the Tidyverse, open R or RStudio and type `install.packages("tidyverse")` into the console.

@@

#### **Getting help in R**

@@

1. If you are ever unsure what a function does you can type `?functionname` or `help(functionname)` into the console and a help file will pop up (in the bottom right corner of your screen if you are using R Studio) that has useful information about the function.

@@

2. If you are curious about an entire package you can type vignette(packagename) for more information about the package. Note that not all packages have a vignette.

@@

3. If you need help resolving a more complex problem, try Google or another search engine first. There are many people working in R and many sites which aggregate responses to frequently asked questions. [Stackoverflow](https://stackoverflow.com/) is one that is well-known in the coding community and that I personally use regularly while at work. I encourage you to investigate solutions to your issues yourself first because you will retain more information when you've had to spend some time searching and iterating on your own.

### Helpful Links

@@

## Lesson 2 - Variables, data types, and operators {.tabset .tabset-fade .tabset-pills}

### Topics Covered

@@

### Lesson

#### **Printing output**

@@

Another way that is often used is to wrap the item you want to show in the `print()` function.

@@

#### **An introduction to variables**

Variables in R are convenient ways to keep track of values. There are times when you will want to perform an operation, save the value, perform some intermediate steps, and then use the original value that you saved. Let's work with the example of addition. You could do the following to add 4 twice:

@@

Variables allow you to work with values dynamically and without the struggle of trying to remember all of the values that you are working with individually. Keeping the math theme, suppose you wanted to implement the quadratic formula which is written:<br/>
![](./Lesson 2/quadratic_formula.png)<br/>

@@

#### **Assigning variables in R**

@@

There is nothing wrong with using any of the assignment operators so you should go with the one you are most comfortable with, especially as you learn. The industry standard, however, is to use the left-arrow unless there is good reason. This prevents confusion when you are trying to do a check for equality (`==`) and may be easier to read than the right-arrow. I will use the left-arrow exclusively throughout the remainder of this course.

#### **Basic data types**

@@

1. Logical
    + True/false data type (Boolean).
2. Numeric
    + All real numbers.
3. Integer
    + Real values without decimal points.
4. Complex
    + Imaginary values
5. Character
    + String values. These are values that can contain non-numeric characters (such as letters).
6. Raw
    + Specifies values as raw bytes

@@

##### **Logical**

Data with class logical can only take on the values of true or false. Another name for this type of data is Boolean. In R, `TRUE` and `FALSE` are equivalent to 1 and 0 respectively. That means you can do things like add and subtract logical data even though they are not technically numeric.

@@

##### **Numeric**

@@

```{r l2_num_ex1}
class(4)
```

You might wonder how R can tell the difference between a numeric 4 and an integer 4. Mathematically, after all, 4 is both an integer and a real number. The way to pass an explicitly integer value to R is to append an `L` to the end of the number. For example:

```{r l2_num_ex2}
class(4L)
class(1040L)
```

Appending an `L` will not convert a number which has a decimal into an integer:

```{r l2_num_ex3}
class(1.33L)
```

Finally, there are other functions that can convert from numeric to integer for you (and indeed convert between any of the types):

```{r}
x <- 4
class(x)
x <- as.integer(x)
class(x)
```

##### **Character**

@@

The first line of code throws an error, telling us that "1" and "2" are not numeric and therefore cannot be added. The second line of code returns what you expect, 3. Folks coming from other programming languages should note that string concatenation cannot be done with the `+` operator in R (although you could create an operator that does this yourself).

```{r l2_char_ex3}
`%$$%` <- function(lhs, rhs) {
  out <- paste0(lhs, rhs)
  return(out)
}

"Hel" %$$% "lo"
"How" %$$% " are" %$$% " you?"
```

#### **Mathematical operators**

@@

#### **Logical operators**

You are also likely familiar with most of the logical operators. These are things in math which do not return a number, but rather return a value of `TRUE` or `FALSE`. Greater than (`>`), less than (`<`), greater than or equal to (`>=`), less than or equal to (`<=`), equal to (`==`), and not equal to (`!=`) are all available in R. Note that you must use 2 equal signs in R when you want to equate two things. One equal sign, as we discussed earlier, is for assigning values to variables.

@@

There are also operators for *or* (`|`), *and* (`&`), and *not* (`!`).

@@

### Helpful Links

* [You can read more about the data types that we do not cover here.](https://www.programiz.com/r/data-types)
* [More information about operators](https://www.statmethods.net/management/operators.html)
* [For a deeper dive into creating custom operators](https://adv-r.hadley.nz/functions.html#infix-functions)

## Lesson 3 - Tidy data and dplyr {.tabset .tabset-fade .tabset-pills}

### Topics Covered

@@

### Lesson

#### **Why dplyr?**

@@

#### **%>% and dplyr functions**

##### **%>%**

@@

##### **filter**

@@

##### **slice**

@@

##### **arrange**

@@

##### **select**

@@

##### **rename**

@@

##### **mutate**

@@

##### **relocate**

@@

##### **summarize**

@@

## Lesson 4 - Vectors and lists {.tabset .tabset-fade .tabset-pills}

### Topics Covered

@@

## Lesson 5 - Conditionals and loops {.tabset .tabset-fade .tabset-pills}

### Topics Covered

@@

## Lesson 6 - Functions {.tabset .tabset-fade .tabset-pills}
